\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-16 doc/mlsys_paper/paper.tex
changed ugly footnote url to inline listin; various edits to improve clarity and compactness

@@

% Alphabetical ordering?
% TODO: check affiliations
\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
Yannis Mentekidis
% any affiliation/email?
\And
Conrad Sanderson \\
Data61, CSIRO \\
Australia
}

@@

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:
%
%\vspace*{-0.2em}
\begin{equation}
\operatorname{argmin} f(x)
\end{equation}
\vspace*{-1.5em}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present {\tt ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template
metaprogramming~\cite{Alexandrescu2001},
{\tt ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

@@

{\tt ensmallen} is open-source software licensed under the 3-clause BSD
license\footnote{\url{https://opensource.org/licenses/BSD-3-Clause}},
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations.

@@

{\tt ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\operatorname{argmin}_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for {\tt
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.
%% CS: in the above sentence, it's not clear what is meant by "we must design ..."
%% CS: should it be "we provide ..." or "ensmallen provides ..." ?
%% CS: The overall intent of the sentence is not clear -- to my eyes
%% CS: it doesn't seem to gel well with its surrounding context.

@@

%% CS: i've added \nolimits to save a bit of space
\vspace*{-1.1em}
\begin{equation}
% f(\theta) =   %% CS: for clarity, perhaps we should have f(\theta) = ...
\operatorname{argmin}_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-1.1em}

The above objective function $f(\theta)$ has the associated gradient

@@

We can implement these two functions in a class {\tt LinearRegressionFunction}.
Here {\tt arma::mat} and {\tt arma::vec} represent Armadillo matrix and vector
types, respectively~\cite{sanderson2016armadillo}.

% Indented a little to make it clear it's different from the text...

%% CS: "accu((y - X * theta) % (y - X * theta))" is redoing computations
%% CS: ie. efficiency problem
%% CS: how about:
%% CS:   arma::mat tmp = y - X * theta;
%% CS:   return arma::accu(tmp % tmp);  // accumulate element-wise product
%% CS: in the above code, the comment hints at what the % operator does,
%% CS: as otherwise it's potentially confusing to someone not familiar with Armadillo
\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    class LinearRegressionFunction {
     public:
      // Construct the LinearRegressionFunction with the given data
      LinearRegressionFunction(arma::mat& X, arma::vec& y) : X(X), y(y) {}

      // Compute the objective function
      double Evaluate(const arma::mat& theta) {
        return std::pow(arma::norm(y - X * theta), 2.0);
      }
      // Compute the gradient and store in 'gradient'
      void Gradient(const arma::mat& theta, arma::mat& gradient) {
        gradient = -2 * X.t() * (y - X * theta);
      }
     private:
      arma::mat& X; arma::vec& y;
    };
\end{minted}
\vspace*{-0.4em}
%% CS: the above code is currently shown across two pages - not a clean look.
%% CS: how about we place it in a figure ?

This is the entire required implementation to optimize the linear regression model with
any of {\tt ensmallen}'s gradient-based optimizers.  The code snippet below
shows how the L-BFGS optimizer can be used to find the best parameters $\theta$:

@@

\vspace*{-0.3em}
\section{Template Metaprogramming for Ease of Use and Efficiency}
\vspace*{-0.5em}

When optimizing a given function $f(x)$, the computation of
the objective function $f(x)$ and its derivative $f'(x)$ often involve the
computation of identical intermediate results.  Consider the linear regression
objective function described
in Section~\ref{sec:linreg_example}, $f(\theta) = \| y - X\theta \|_F^2$.
For this objective function, the derivative $f'(x)$ has a related form of
$-2 X^T (y -X \theta)$.  Both the objective function and the derivative 
depend on the computation of the vector term $(y - X \theta)$,
which can be computationally expensive to compute if $X$ is a large matrix.
Existing optimization frameworks do not have an easy way to avoid
this duplicate computation. In many cases, an optimization algorithm
may need the values of both $f(\theta)$ and $f'(\theta)$ for a given $\theta$.

Using template metaprogramming, {\tt ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  For the 
\texttt{LinearRegressionFunction} in Section~\ref{sec:linreg_example},
we can use an implementation that only computes $(y - X \theta)$ once:

\vspace*{-0.5em}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::rowvec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\vspace*{-0.5em}

@@

This method can be provided {\it instead} of {\tt Evaluate()} and {\tt
Gradient()}.  In either case, template metaprogramming techniques are used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide any `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if {\tt EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both {\tt Gradient()} and {\tt Evaluate()} in turn.
Similarly, if {\tt Evaluate()} or {\tt Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

The use of templates in this manner also allows for compiler optimizations that
would not otherwise be possible (and that are often not possible in other
frameworks).  Firstly, because the objective function class itself is a template
parameter, the compiler is able to avoid the overhead of a function pointer
dereference, which would not be easily possible when using a language with
virtual inheritance.  The compiler is also able to use inlining and any
optimizations that may imply, including removing temporary value and dead code
elimination.  Further, if {\tt ensmallen} automatically generates an {\tt
Evaluate()} or {\tt Gradient()} method from a user-supplied {\tt
EvaluateWithGradient()} method, the compiler can in some cases recognize and
remove the computation of unnecessary results.  For instance, in an
automatically generated {\tt Evaluate()} method, the computation of the gradient
from {\tt EvaluateWithGradient()} can be avoided entirely.

@@

To demonstrate the benefits of these optimizations that {\tt ensmallen} can take
advantage of, we compare {\tt ensmallen}'s performance to a number of other
optimization frameworks, including some that use automatic differentiation.

For our first experiment, we consider the simple and popular Rosenbrock
function~\cite{Rosenbrock1960}: $f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 -
x_1^2)$.  For the optimizer, we use simulated
annealing~\cite{kirkpatrick1983optimization}, a gradient-free optimizer.
Simulated annealing will call the objective function very many times; for each
simulation we limit the optimizer to 100k objective evaluations.  Since the
objective function is so simple and is called so many times, this can help us
understand the overheads of different frameworks.

In our comparisons we use four different frameworks:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} from GNU Octave's {\tt optim} package~\cite{octave}
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}
\vspace*{-0.3em}

@@

Initially, we implemented these functions as simply as possible and ran them
without any tuning, which reflects how a normal user might interact with the
software.  These results are seen in the first row of
Table~\ref{tab:rosenbrock_results}.  Of the four frameworks, only Julia and {\tt
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

@@

Next, we consider the same machine learning example we have been working with
through the paper: linear regression.  For this task we will use the first-order
L-BFGS optimizer~\cite{zhu1997algorithm},
so for {\tt ensmallen} we will implement two versions: one with {\tt Evaluate()}
and {\tt Gradient()}, and one with only {\tt EvaluateWithGradient()}.  The code
for each of these is the same as shown earlier.  In addition, we have more
options for Julia: we can implement only the objective function and allow either
Julia's {\tt
Calculus.jl}\footnote{\url{https://github.com/JuliaMath/Calculus.jl}} or {\tt
ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} package to automatically
compute the gradient.  Therefore, we compare against the following:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses GNU Octave's {\tt optim} package
\end{itemize}
\vspace*{-0.3em}

Results for different data sizes are shown in Table~\ref{tab:lbfgs}.  For each
implementation, L-BFGS was allowed to run for only $10$ iterations and never
converged in fewer iterations.  The datasets trained on were highly noisy random
data with a slight linear pattern---but note that the exact data is not relevant
for our experiments here, only its size.  Runtimes are again reported as the
average across 10 runs.

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen} & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with several different dataset
sizes.  All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

We can see that the use of {\tt EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the {\tt ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for the user to write their
code, it is very clear that the code generated by the automatic differentiator
is far from optimal, with runs that take up to three orders of magnitude longer!

>>>>

% \DeclareMathOperator requires \usepackage{amsmath}
\DeclareMathOperator*{\argmindown}{argmin}   % limit is placed underneath
\DeclareMathOperator{\argminright}{argmin}   % limit is placed on the right side
% more details:
% https://tex.stackexchange.com/questions/5223/command-for-argmin-or-argmax

@@

% Alphabetical ordering?
% TODO: check affiliations
\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
%% CS: I've added "Independent Researcher" below for now,
%% CS: as a blank affiliation looks weird and incomplete
Yannis Mentekidis \\
Independent Researcher \\
\texttt{mentekid@gmail.com}
% any affiliation/email?
%% CS: googling suggests that Yannis is/was affiliated with Aristotle University of Thessaloniki 
%% CS: and is perhaps now with Amazon
\And
%% CS: I have two affiliations, so I've listed them on two lines
Conrad Sanderson \\
Data61, CSIRO, Australia \\
University of Queensland, Australia\\
\texttt{conradsand@ieee.org}
}

@@

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:
%
%\vspace*{-0.2em}
\begin{equation}
\argmindown_x f(x)
\end{equation}
\vspace*{-1.5em}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present {\tt ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template
metaprogramming~\cite{Abrahams_2004},
{\tt ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

@@

{\tt ensmallen} is open-source software licensed under the 3-clause BSD
license~\cite{Rosen_2004_full}, 
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations.

@@

{\tt ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\argminright_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for {\tt
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.
%% CS: in the above sentence, it's not clear what is meant by "we must design ..."
%% CS: should it be "we provide ..." or "ensmallen provides ..." ?
%% CS: The overall intent of the sentence is not clear -- to my eyes
%% CS: it doesn't seem to gel well with its surrounding context.

@@

%% CS: i've added \nolimits to save a bit of space
\vspace*{-1.1em}
\begin{equation}
\argmindown_\theta f(\theta) =   %% CS: for clarity
\argmindown_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\argmindown_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-1.1em}

The objective function $f(\theta)$ has the associated gradient

@@

We can implement these two functions in a class named {\tt LinearRegressionFunction},
as shown in Fig.~\ref{fig:LinearRegressionFunction}.
This is the entire required implementation to optimize the linear regression model with
any of the gradient-based optimizers in {\tt ensmallen}.

\begin{figure}[!tb]
\hrule\vspace*{0.3ex}
\begin{minted}[fontsize=\small]{c++}
class LinearRegressionFunction {
 public:
  // Construct the LinearRegressionFunction with the given data
  LinearRegressionFunction(arma::mat& X, arma::vec& y) : X(X), y(y) {}

  // Compute the objective function
  double Evaluate(const arma::mat& theta) {
    return std::pow(arma::norm(y - X * theta), 2.0);
  }
  // Compute the gradient and store in 'gradient'
  void Gradient(const arma::mat& theta, arma::mat& gradient) {
    gradient = -2 * X.t() * (y - X * theta);
  }
 private:
  arma::mat& X; arma::vec& y;
};
\end{minted}
\hrule\vspace*{0.3ex}
\caption
  {
  Implementation of objective and gradient functions for linear regression,
  to be used by optimizers in \texttt{ensmallen}.
  The types {\tt arma::mat} and {\tt arma::vec} are 
  matrix and vector types from Armadillo~\cite{sanderson2016armadillo}.
  }
\label{fig:LinearRegressionFunction}
\end{figure}

Given the user-defined {\tt LinearRegressionFunction} class,
the code snippet below
shows how the L-BFGS optimizer can be used to find the best parameters $\theta$:

@@

\vspace*{-0.3em}
\section{Automatic Metaprogramming for Ease of Use and Efficiency}
\vspace*{-0.5em}

When optimizing a given function $f(x)$, the computation of
the objective function $f(x)$ and its derivative $f'(x)$ often involve the
computation of identical intermediate results.  Consider the linear regression
objective function described
in Section~\ref{sec:linreg_example}, $f(\theta) = \| y - X\theta \|_F^2$.
For this objective function, the derivative $f'(\theta)$ has a related form of
$-2 X^T (y -X \theta)$.  Both the objective function and the derivative 
depend on the computation of the vector term $(y - X \theta)$,
which can be computationally expensive to compute if $X$ is a large matrix.
Existing optimization frameworks do not have an easy way to avoid
this duplicate computation. In many cases, an optimization algorithm
may need the values of both $f(\theta)$ and $f'(\theta)$ for a given $\theta$.

Using template metaprogramming, {\tt ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  As an example,
for the 
\texttt{LinearRegressionFunction} class in Fig.~\ref{fig:LinearRegressionFunction},
we can replace {\tt Evaluate()} and {\tt Gradient()}
with an implementation of {\tt EvaluateWithGradient()}
that only computes $(y - X \theta)$ once:

\vspace*{-0.5em}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::rowvec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;  // Store gradient in the provided matrix
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\vspace*{-0.5em}

@@

Template metaprogramming techniques are automatically used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide any `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if {\tt EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both {\tt Gradient()} and {\tt Evaluate()} in turn.
Similarly, if {\tt Evaluate()} or {\tt Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

The use of template metaprogramming in this manner also allows for compiler optimizations that
would not otherwise be possible (and that are often not possible in other
frameworks).  Firstly, because the objective function class itself is a template
parameter, the compiler is able to avoid the overhead of a function pointer
dereference, which would not be easily possible when using a language with
virtual inheritance.  The compiler is also able to use inlining and any
optimizations that may imply, including removing temporary values and dead code
elimination.  Further, if {\tt ensmallen} automatically generates an {\tt
Evaluate()} or {\tt Gradient()} method from a user-supplied {\tt
EvaluateWithGradient()} method, the compiler can in some cases recognize and
remove the computation of unnecessary results.  For instance, in an
automatically generated {\tt Evaluate()} method, the computation of the gradient
from {\tt EvaluateWithGradient()} can be avoided entirely.

@@

To demonstrate the benefits of the metaprogramming based code optimizations
that {\tt ensmallen} can exploit,
we compare the performance of {\tt ensmallen} with several other
optimization frameworks, including some that use automatic differentiation.

For our first experiment, we consider the simple and popular Rosenbrock
function~\cite{Rosenbrock1960}: $f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 -
x_1^2)$.  For the optimizer, we use simulated
annealing~\cite{kirkpatrick1983optimization}, a gradient-free optimizer.
Simulated annealing will call the objective function numerous times; for each
simulation we limit the optimizer to 100k objective evaluations.  Since the
objective function is straightforward and is called many times, this can help us
understand the overheads of various frameworks.  We compare four frameworks%
%
\footnote{Another option here might be {\tt simulannealbnd()} 
in the Global Optimization Toolkit for MATLAB.
However, no license was available for these simulations.}
%
for this task:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item simulated annealing implementation in {\tt Optim.jl} with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} in the {\tt optim} package for GNU Octave~\cite{octave}
\end{itemize}
\vspace*{-0.3em}

@@

Initially, we implemented these functions as simply as possible and ran them
without any tuning, which reflects how a normal user might interact with the
software.  These results are seen in the first row of
Table~\ref{tab:rosenbrock_results}.  Only Julia and {\tt
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

%% CS: i'm not convinced that this conjecture is barking up the right tree;
%% CS: SciPy is in Python, which is by default an interpreted language.
%% CS: the slowdown many have nothing to do with inlining and simply due to
%% CS: Python being slow in general.
%% CS: it's possible to use PyPy or Numba to speed things up,
%% CS: in order to provide a "fairer" comparison against compiled languages.
%% CS: Octave also has an optional JIT compiler, which last time I checked
%% CS: needs to be explicitly enabled.
%% 
%% CS: i would suggest to stay away from the "we do inlining" argument,
%% CS: as it can be blown up too easily.
%% CS: furthermore, I would suggest to mention PyPy and/or Numba and Octave JIT
%% CS: (in order to show that we've done our homework),
%% CS: and then state something along the lines of
%% CS: "these approaches are not being used as we're comparing out-of-the-box performance"

% RC: agree, hopefully the paragraph below pushes this in more of a right
% direction.

@@

Next, we consider the linear regression example described in Sec.~\ref{sec:linreg_example}.
For this task we use the first-order L-BFGS optimizer~\cite{zhu1997algorithm}.
For {\tt ensmallen} we have 2 versions:
(i)~with only {\tt EvaluateWithGradient()},
and
(ii)~with {\tt Evaluate()} and {\tt Gradient()}.
The code for these functions is as shown earlier.
For Julia we have the options of using manually defined objective and gradient functions,
or the gradient function can be automatically computed by 
{\tt Calculus.jl}
({\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl})
or {\tt ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016}.
Overall we compare across the following:

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}-1: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{itemize}
\vspace*{-0.3em}

Results for various data sizes are shown in Table~\ref{tab:lbfgs}.  For each
implementation, L-BFGS was allowed to run for only $10$ iterations and never
converged in fewer iterations.  The datasets used for training are highly noisy random
data with a slight linear pattern. Note that the exact data is not relevant
for the experiments here, only its size.  Runtimes are reported as the
average across 10 runs.

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen}-1 & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

The results indicate that {\tt ensmallen} with {\tt EvaluateWithGradient()}
is the fastest approach.
Furthremore, the use of {\tt EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the {\tt ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for the user to write their
code, it is clear that the code generated by the automatic differentiators
({\tt Calculus.jl} and {\tt ForwardDiff.jl})
is far from optimal, with runs that are slower by up to three orders of magnitude.

% TODO: show flexibility of optimization with learning curves:
%  - use LinearRegressionFunction modified for small batches
%  - make sure Info or Debug output is on
%  - run with a whole boatload of SGD variants
%  - parse the output with awk/sed into a csv of objectives per epoch
%  - plot it
%  - profit!
%
% Probably a snippet showing the actual code to run with a bunch of different
% optimizers is good too.  Other things can be cut to make space.
